Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-10-05 main.tex
reread

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter         & Range \\
  \hline
  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thr} \}$ \\
  $r$ (ratio)        & $(0,1)$ \\
  eps (epsilon)            & $[0,4]$ \\
  thresh (threshold)          & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
\begin{align}
  \label{eq:r}
 r & = \frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.
\end{align}
Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} & = 
  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
  & = n_{\text{treat}}\cdot
  \frac{r}{k_1\cdot
  r+k_2\cdot (1-r)}, \nonumber \\
  n_{\text{stage2}} & =
  n_{\text{treat}}\cdot
  \frac{1-r}{k_1\cdot
  r+k_2\cdot (1-r)}.
  %\end{split}
\end{align}
%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
This is only the case for selection strategies 1, 2, and 3.
For the selection strategies \emph{threshold} and \emph{epsilon}, $k_2$ depends on the outcome of stage 1.
%choice of the \emph{thresh} or \emph{eps} parameter.
In these cases, for given values of $n_{\text{stage1}}$ and \emph{thresh} or \emph{eps}, respectively, we can  obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed values for $n_{\text{stage1}}$.
Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
More precisely, we minimize
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  h(n_{\text{stage1}}) &:= \left( \Big(k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot %\underbrace{
  \frac{1-r}{r} \cdot n_{\text{stage1}}
  %}_{=n_{\text{stage2}}}
  \Big) - n_{\text{treat}} \right)^2
%  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
with respect to $n_{\text{stage1}}$, where the estimate $\hat{k}_2$ is obtained as described above dependent on $n_{\text{stage1}}$. 
%$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
%This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
%The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.

The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{thr}$ and $\text{eps}$. Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
\begin{equation}
  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
\end{equation}
In our simulations, $n_{\text{treat}}$ is considered to be a constant.

>>>>

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than \emph{eps} (epsilon) below maximal test statistic \\
  \emph{thresh}     & Select all arms with test statistic larger than the value \emph{thresh} (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter         & Range \\
  \hline
  Selection strategy & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
  $r$ (ratio)        & $(0,1)$ \\
  eps (epsilon)            & $[0,4]$ \\
  thresh (threshold)          & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

%FIXME: JR: uniform naming: eps, thersh for the values and epsilon and treshold for the strategies

@@

The total number of treatments $n_{\text{treat}}$ should be constant in the simulation study, when comparing the efficiency of different designs.
We denote the proportion of patients per treatment in stage 1, compared to patients per treatment in both stages, with 
\begin{align}
  \label{eq:r}
 r & = \frac{n_{\text{stage1}}}{n_{\text{stage1}}+n_{\text{stage2}}}.
\end{align}
Given the number of treatments $k_1$ and $k_2$ in the two stages and the total number of patients $n_{\text{treat}}$, optimizing the power with respect to $n_{\text{stage1}}$ and $n_{\text{stage2}}$ is equivalent to optimizing with respect to $r$, since
\begin{align}
  \label{eq:stagec}
  %\begin{split}
  n_{\text{stage1}} & = 
  n_{\text{stage1}}\cdot \frac{n_{\text{treat}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} = 
  n_{\text{treat}} \cdot \frac{n_{\text{stage1}}}{k_1\cdot n_{\text{stage1}}+k_2\cdot n_{\text{stage2}}} \nonumber \\
  & = n_{\text{treat}}\cdot
  \frac{r}{k_1\cdot
  r+k_2\cdot (1-r)}, \nonumber \\
  n_{\text{stage2}} & =
  n_{\text{treat}}\cdot
  \frac{1-r}{k_1\cdot
  r+k_2\cdot (1-r)}.
  %\end{split}
\end{align}
%For $k_1 = k_2 = 5$, $c$ would be $n_{\text{treat}} \cdot \frac{1}{5}$ because for each arm we can distribute $\frac{1}{5}$ of the treatments across stage 1 and stage 2.
These formulas give the numbers of patients per treatment arm in the two stages, if $k_2$ is known. 
This is only the case for selection strategies 1, 2, and 3.
For the selection strategies \emph{threshold} and \emph{epsilon}, $k_2$ depends on the outcome of stage 1.
%choice of the \emph{thresh} or \emph{eps} parameter.
In these cases, for given values of $n_{\text{stage1}}$ and \emph{thresh} or \emph{eps}, respectively, we can  obtain an estimate $\hat{k}_2$ of $k_2$ with a so-called calibration step.
For a set of values for $n_{\text{stage1}}$ in the interval $\left[ \lceil 0.01 \cdot \frac{n_{\text{treat}}}{k_1} \rceil, \lceil \frac{n_{\text{treat}}}{k_1} \rceil \right]$, we simulate stage 1 of the trial multiple times.
We calculate $n_{\text{stage2}} = (1-r)/r\cdot n_{\text{stage1}}$ from formula (\ref{eq:r}) and $\hat{k}_2$ as the average of the resulting values of $k_2$ for all simulations with a fixed value for $n_{\text{stage1}}$.
Then we select the value of $n_{\text{stage1}}$ such that the corresponding number of total patients is closest to $n_{\text{treat}}$.
More precisely, we minimize
\begin{align}
  \label{eq:targettreat}
  \begin{split}
  h(n_{\text{stage1}}) &:= \left( \Big(k_1 \cdot n_{\text{stage1}} + \hat{k}_2 \cdot %\underbrace{
  \frac{1-r}{r} \cdot n_{\text{stage1}}
  %}_{=n_{\text{stage2}}}
  \Big) - n_{\text{treat}} \right)^2
%  \hat{k}_2 &= f(n_{\text{stage1}}, 1, \text{effect set}, \text{Selection strategy}, \text{Epsilon}, \text{Threshold})_{k_2}
  \end{split}
\end{align}
with respect to $n_{\text{stage1}}$, where the estimate $\hat{k}_2$ is obtained as described above dependent on $n_{\text{stage1}}$. 
%$f$ is the simulator black-box from Equation~\ref{eq:bbox}, that gives us the estimated number of $\hat{k}_2$.
%This number is independent of the simulation of the second stage, so we can set $n_{\text{stage2}}$ to 1.
%The value of $\hat{k}_2$ that was observed with the minimal outcome of $h$ is then used to calculate $c$ in equation~\ref{eq:ntreatcc}.

The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{threshold}$ and $\text{epsilon}$. Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
\begin{equation}
  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \text{epsilon}, \text{threshold}).
\end{equation}
In our simulations, $n_{\text{treat}}$ is considered to be a constant.
